---
title: "Find the optimum number of clusters for hierarchical clustering"
author: "Sehyun Oh"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: false
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})
```

# Preparae negative controls
## Build negative controls
Source the below script will execute the following processes:    

1. Generate 50 synthetic datasets:   
   - Each dataset contains 50 random samples from 44,890 samples (with replacement).   
   - Scaramble genes in each dataset by random selection without replacement and
   add random value between -0.1 and 0.1   
   
2. Row-normalize synthetic datasets   

3. PCA on synthetic datasets --> `bootstrap_PCs_rowNorm_Neg.rds`   

4. Combine top 20 PCs from traning datasets and PC1s from synthetic datasets --> `all_{#neg}.rds`   

5. Calculate distance matrix and hcut for each combined dataset --> `res_dist_{#neg}.rds` and `res_hclust_{#neg}.rds`   

6. Evaluate how the negative controls were separated --> `evals_{#neg}.rds` and `eval_summary_{#neg}.rds`   

```{r eval=FALSE}
## Do not run this script without checking it first!!
source("neg_controls.R")
```


## Validate negative controls
Below is the summary of the distance matrix of 10,720 PCs (top 20 PCs from 536 studies).
We want to create a synthetic dataset as a negative control with the distance 
rangeing median or higher.

```{r collapse=FALSE}
wd <- "~/data2/PCAGenomicSignatureLibrary/refinebioRseq/PCAmodel_536"
res_dist <- readRDS(file.path(wd, "res_dist.rds"))
summary(res_dist)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 0.01688 0.98783 1.00000 0.99999 1.01217 1.97482
```

We tested the distance matrix with 50 'negative controls', generated from the 
`neg_controls.R` script above. Distance ranges roughly between 1st quarter and 3rd 
quarter of training datasets' distance distribution. 

```{r collapse=FALSE}
dat_dir <- "~/data2/PCAGenomicSignaturesPaper/inst/extdata/Neg_Controls"
neg <- readRDS(file.path(dat_dir, "bootstrap_PCs_rowNorm_Neg.rds"))
data <- lapply(neg, function(x) x$rotation) %>% Reduce(cbind,.) %>% t

ind <- c()
for (i in 1:50) {new_ind = c(1) + 20*(i-1); ind = c(ind, new_ind)} # only PC1s
control_dat <- data[ind,]

res.dist.neg <- factoextra::get_dist(control_dat, method = "spearman")
summary(res.dist.neg)
```

We plot the heatmap of distance matrix with different number of negative controls
(PC1s from synthetic datasets) and confirmed that they can serve as a negative control.

```{r fig.width=6, fig.height=5.5, echo=FALSE}
dat_dir <- "~/data2/PCAGenomicSignaturesPaper/inst/extdata/Neg_Controls"
neg <- readRDS(file.path(dat_dir, "bootstrap_PCs_rowNorm_Neg.rds"))
data <- lapply(neg, function(x) x$rotation) %>% Reduce(cbind,.) %>% t

dist_med <- 1.00000
dist_max <- 1.97482

numOfDatasets <- c(10, 20, 30, 40, 50)
numOfTopPCs <- 20

for (numOfDataset in numOfDatasets) {
  ind <- c()
  for (i in 1:numOfDataset) {new_ind = c(1) + numOfTopPCs*(i-1); ind = c(ind, new_ind)}
  control_dat <- data[ind,]

  res.dist.neg <- factoextra::get_dist(control_dat, method = "spearman")
  plot <- ComplexHeatmap::Heatmap(as.matrix(res.dist.neg),
                                  row_names_gp = grid::gpar(fontsize = 5),
                                  column_names_gp = grid::gpar(fontsize = 5),
                                  column_title = "Distance Method = spearman",
                                  column_title_side = "top",
                                  col = circlize::colorRamp2(breaks = c(0, dist_med, dist_max), 
                                                  colors = c("blue", "white", "red")))
  print(plot)
}
```

```{r save_heatmap50, echo=FALSE}
plot <- ComplexHeatmap::Heatmap(as.matrix(res.dist.neg),
                                row_names_gp = grid::gpar(fontsize = 5),
                                column_names_gp = grid::gpar(fontsize = 5),
                                column_title = "A. Distance Matrix",
                                column_title_side = "top",
                                col = circlize::colorRamp2(breaks = c(0, dist_med, dist_max), 
                                                colors = c("blue", "white", "red")))

png("outputs/png/A.heatmap_neg50.png", width = 400, height = 380)
plot
dev.off()
```



# Optimum k for different numbers of controls
Next we combined top 20 PCs of training datasets and PC1s from varying numbers of 
negative controls (10, 20, 30, 40, and 50) and performed hierarchical clustering 
with different numbers of clusters. Here, we plotted the proportion of separated 
negative controls against the number of clusters.

Number of clusters for each combined datastes were rounds of the number of PCs 
divided by 9 different numbers (7, 6, 5, 4, 3, 2.75, 2.5, 2.25, and 2).

```{r}
dat_dir <- "~/data2/PCAGenomicSignaturesPaper/inst/extdata/Neg_Controls"
numOfDatasets <- c(10, 20, 30, 40, 50)

## Load all eval_summary
eval_res <- list()
for (i in numOfDatasets) {
  fname <- paste0("eval_summary_", i)
  res <- readRDS(file.path(dat_dir, paste0(fname, ".rds"))) %>% as.data.frame 
  eval_res[[fname]] <- res
}

## Combine all eval_summary
df <- Reduce(rbind, eval_res) 
df$numOfControls <- rep(numOfDatasets, each = 9) %>% as.character   # length(k_range) = 9
df$separtedPortion <- as.numeric(df$numSeparated)/as.numeric(df$numOfControls)
df$numUnseparated <- as.numeric(df$numOfControls) - as.numeric(df$numSeparated)
```

## Proportion of separated negative controls
```{r plot1, echo=FALSE, fig.width=6.5, fig.height=4}
plot1 <- ggplot(df, aes(x = numOfCluster, y = separtedPortion)) + 
  geom_line(aes(colour = numOfControls)) + 
  geom_point() + 
  ggtitle("B. Proportion of separated negative controls")

plot1

## Save plot1
png("outputs/png/B.separatedPortion.png", width = 500, height = 350)
plot1
dev.off()
```

## Number of non-separated negative controls
```{r plot2, echo=FALSE, fig.width=6.5, fig.height=4}
df$numUnseparated <- as.numeric(df$numOfControls) - as.numeric(df$numSeparated)
plot2 <- ggplot(df, aes(x = numOfCluster, y = numUnseparated)) + 
  geom_line(aes(colour = numOfControls)) + 
  geom_point() + 
  ggtitle("C. Number of non-separated negative controls")

plot2

## Save plot2
png("outputs/png/C.numUnseparated.png", width = 500, height = 350)
plot2
dev.off()
```

```{r}
sessionInfo()
```


