---
title: "Reproduce multiPLIER Figure 5 - Pathway Separation"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>", message = FALSE, warning = FALSE, collapse = FALSE
)
```

# Setup
We benchmarked the model validation measure, named as pathway separation, Figure 5 from the 
previous [study](https://www.sciencedirect.com/science/article/pii/S240547121930119X?via%3Dihub), 
where authors expanded the [PLIER](https://www.nature.com/articles/s41592-019-0456-1) 
method to the recount2 dataset. Briefly, pathway separation is defined as the ability 
of the signature model to keep non-overlapping signatures that can differentiate 
biologically similar pathways. 

```{r}
suppressPackageStartupMessages({
  library(PCAGenomicSignaturesPaper)
  library(PCAGenomicSignatures)
  library(dplyr)
})
```


## PCAmodel
To directly compare with the results from multiPLIER paper, we used the PCAmodel
annotated with the same priors: bloodCellMarkersIRISDMAP, svmMarkers, and canonicalPathways.

```{r message=FALSE, warning=FALSE}
data.dir <- system.file("extdata", package = "PCAGenomicSignaturesPaper")
PCAmodel <- readRDS(file.path(data.dir, "PCAmodel_PLIERpriors.rds"))
PCAmodel
updateNote(PCAmodel)
```



# Pathway Separation
## Type I and type II interferon
`cutoff_n` argument of `checkPathwaySeparation` function decides how many of the 
top enriched pathways are used for comparison. We tried both top 5 and only the 
top 1. 

```{r}
ifn.alpha.set <- c("REACTOME_INTERFERON_ALPHA_BETA_SIGNALING")
ifn.gamma.set <- c("REACTOME_INTERFERON_GAMMA_SIGNALING")

checkPathwaySeparation(PCAmodel, ifn.alpha.set, ifn.gamma.set, 
                       cutoff_nes = NULL, cutoff_n = 5)
checkPathwaySeparation(PCAmodel, ifn.alpha.set, ifn.gamma.set, cutoff_n = 1)
```


## Myeloid lineage

```{r}
neutrophil.set <- c("DMAP_GRAN3", "IRIS_Neutrophil-Resting", "SVM Neutrophils")
monocyte.set <- c("IRIS_Monocyte-Day0", "IRIS_Monocyte-Day1", 
                  "IRIS_Monocyte-Day7", "DMAP_MONO2", "SVM Monocytes",
                  "SVM Macrophages M0", "SVM Macrophages M1", 
                  "SVM Macrophages M2")

checkPathwaySeparation(PCAmodel, neutrophil.set, monocyte.set, 
                       cutoff_nes = NULL, cutoff_n = 5) 
checkPathwaySeparation(PCAmodel, neutrophil.set, monocyte.set, cutoff_n = 1)
```


## Proliferation
Molecular processes we would associate with proliferating cells, namely the
G1 and G2 phases of the cell cycle.

```{r}
g1.set <- c("REACTOME_G1_S_TRANSITION", "REACTOME_M_G1_TRANSITION",
            "REACTOME_APC_C_CDH1_MEDIATED_DEGRADATION_OF_CDC20_AND_OTHER_APC_C_CDH1_TARGETED_PROTEINS_IN_LATE_MITOSIS_EARLY_G1", 
            "REACTOME_CYCLIN_E_ASSOCIATED_EVENTS_DURING_G1_S_TRANSITION_", 
            "REACTOME_G1_PHASE", "REACTOME_MITOTIC_M_M_G1_PHASES",
            "REACTOME_P53_DEPENDENT_G1_DNA_DAMAGE_RESPONSE", 
            "REACTOME_MITOTIC_G1_G1_S_PHASES", 
            "REACTOME_P53_INDEPENDENT_G1_S_DNA_DAMAGE_CHECKPOINT")
g2.set <- c("REACTOME_MITOTIC_G2_G2_M_PHASES", "REACTOME_G2_M_CHECKPOINTS")

checkPathwaySeparation(PCAmodel, g1.set, g2.set, 
                       cutoff_nes = NULL, cutoff_n = 5) 
checkPathwaySeparation(PCAmodel, g1.set, g2.set, cutoff_n = 1)
```

