---
title: "Find the signature using metadata"
author: "Sehyun Oh"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: false
    toc_depth: 3
  BiocStyle::pdf_document:
    toc: true
    toc_float: false
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r echo=FALSE}
## Packages in GitHub
pkgs <- c("GenomicSuperSignature", "GenomicSuperSignaturePaper")
new_pkgs <- pkgs[!(pkgs %in% rownames(installed.packages()))]
if(length(new_pkgs) != 0) devtools::install_github(paste0("shbrief/", new_pkgs))

## BioC or CRAN packages
pkgs <- c("ggplot2", "ggpubr", "dplyr", "Biobase")
new_pkgs <- pkgs[!(pkgs %in% rownames(installed.packages()))]
if(length(new_pkgs) != 0) BiocManager::install(new_pkgs)

## Load packages
suppressPackageStartupMessages({
  library(GenomicSuperSignature)
  library(GenomicSuperSignaturePaper)
  library(ggplot2)
  library(ggpubr)
  library(dplyr)
  library(Biobase)
})
```

# Setup
## RAVmodel
```{r message=FALSE, warning=FALSE, echo=FALSE}
data.dir <- system.file("extdata", package = "GenomicSuperSignaturePaper")
RAVmodel <- readRDS(file.path(data.dir, "RAVmodel_C2.rds"))
RAVmodel
updateNote(RAVmodel)
```

## CRC validation datasets
```{r}
load("data/eSets/setNames.RData")
```

```{r echo=FALSE, eval=FALSE}
## CRC paper actually used both training and validation datasets (total of 18)
## for Figure 4. So here, we are using all of them. However, for Figure 4C, we'll
## try to use only 10 validation datasets. 
# load(file.path(wd, "trainingSetNames.RData"))
# validationSetNames <- setdiff(setNames, trainingSetNames)
```

```{r}
## Load validation samples
for (set in setNames) {
  load(paste0("data/eSets_new/", set, '.RData'))
}
```



# Processing datasets
## Get phenotype data
Combine all the phenotype data from CRC validation datasets.
```{r}
## phenotype tables combined
pdata_df <- setNames %>% lapply(function(set) {
  eSet <- get(set)
  pdata <- pData(eSet)
  eSet_tmp <- eSet[, pdata$sample_type %in% "tumor"]
  pdata_tmp <- pData(eSet_tmp)
  
  ind_rm <- grep("CRIS_", colnames(pdata_tmp))
  if (length(ind_rm) != 0) {pdata_tmp <- pdata_tmp[,-ind_rm]}
  return(pdata_tmp)
}) %>% Reduce('rbind', .)
```

## Get expression data
Combine all expression profiles from CRC validation datasets, subset with the common
genes among them.

```{r}
## common genes between all validation datasets
all_genes <- list()
for (set in setNames) {
  eSet <- get(set)
  exprs <- exprs(eSet) %>% rmNaInf
  all_genes[[set]] <- rownames(exprs)
}
cg <- Reduce(intersect, all_genes)

## expression matrix combined
exprs_df <- setNames %>% lapply(function(set) {
  eSet <- get(set)
  pdata <- pData(eSet)
  eSet_tmp <- eSet[cg, pdata$sample_type %in% "tumor"]
  exprs_tmp <- exprs(eSet_tmp) %>% rmNaInf
  exprs_tmp <- apply(exprs_tmp, 1, function(x) x - mean(x)) %>% t
  return(exprs_tmp)  
}) %>% Reduce('cbind', .)   # 8219 genes x 3567 samples
```

## Calculate Score
Calculate score and combine it with phenotype data.
```{r}
sampleScore <- calculateScore(exprs_df, RAVmodel)
data_all <- cbind(sampleScore, pdata_df)
```

```{r rm_duplicated, echo=FALSE}
## I added the scores from a few major RAVs to a pData. So remove duplicates.
rm_ind <- which(duplicated(colnames(data_all)))
data_all <- data_all[, -rm_ind]
```




# Multivariable
Here, I'm using 4 CMS subtypes of CRC validation datasets as an example of identifying 
signifcant RAVs using discrete, multivariable metdata.

## ANOVA
```{r ANOVA}
f.stat.all <- sapply(seq_len(ncol(RAVmodel)), function(x) {
  res.aov <- aov(data_all[,x] ~ data_all$cms_label_SSP)
  f.stat <- summary(res.aov)[[1]][1,4]   # extract F-statistics from ANOVA 
  return(f.stat)
})

names(f.stat.all) <- paste0("RAV", seq_len(ncol(RAVmodel)))
head(f.stat.all[order(f.stat.all, decreasing = TRUE)])
```

## Kruskal-Wallis Rank Sum Test
Normality assumption is not met. Below example is a Q-Q plot from RAV834.
```{r fig.width=4, fig.height=4}
res.aov <- aov(data_all[,"RAV834"] ~ data_all[,"cms_label_SSP"])
plot(res.aov, 2)
```

```{r normality_test, echo=FALSE, eval=FALSE}
## Test normality using Shapiro.test :
## [Wikipedia] The null-hypothesis of this test is that the population is normally 
## distributed. Thus, if the p value is less than the chosen alpha level, then the 
## null hypothesis is rejected and there is evidence that the data tested are not 
## normally distributed. 

## Shapiro.test is too sensitive to normality. For example, even if the sample size
## is large enough to ignore normality, Shapiro.test can alarm the normality issue 
## with really high power (with large dataset). Just use Q-Q plot.

normality <- sapply(seq_len(ncol(RAVmodel)), function(x) {
  res.aov <- aov(data_all[,x] ~ data_all$cms_label_SSP)
  aov_residuals <- residuals(object = res.aov)  # Extract the residuals
  res <- shapiro.test(x = aov_residuals)$p.value  # Run Shapiro-Wilk test
  return(res)
})

names(normality) <- paste0("RAV", seq_len(ncol(RAVmodel)))
summary(normality)
```

Try Kruskal-Wallis Rank Sum Test
```{r Kruskal-Wallis}
# Kruskal-Wallis Rank Sum Test
# a non-parametric alternative to one-way ANOVA = when normality assumption is not met
kw.chi.sqr <- sapply(seq_len(ncol(RAVmodel)), function(x) {
  kw.test <- kruskal.test(data_all[,x] ~ data_all$cms_label_SSP)
  kw.stat <- kw.test$statistic
  return(kw.stat)
})

names(kw.chi.sqr) <- paste0("RAV", seq_len(ncol(RAVmodel)))
head(kw.chi.sqr[order(kw.chi.sqr, decreasing = TRUE)])
```

## CMS-associated RAVs
Top 2 RAVs (834 and 833) are same from both ANOVA and Kruskal-Wallis Rank Sum Test.

```{r message=FALSE, warning=FALSE, fig.width=6, fig.height=5.5}
sampleScore1 <- 834
sampleScore2 <- 833
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

Plots with different mean comparing methods.
```{r collapse=FALSE, fig.width=5, fig.height=4}
## t.test for a few pairs
my_comparisons <- list(c("CMS1", "CMS2"),c("CMS1", "CMS3"),c("CMS1", "CMS4"))
ggboxplot(data_all, x = "cms_label_SSP", y = "RAV834", fill = "cms_label_SSP") +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", aes(label = ..p.adj..))

## ANOVA
ggboxplot(data_all, x = "cms_label_SSP", y = "RAV834", fill = "cms_label_SSP") +
  stat_compare_means(method = "anova", label.y = 7) # Add global p-value

## Kruskal-Wallis Rank Sum Test 
ggboxplot(data_all, x = "cms_label_SSP", y = "RAV834", fill = "cms_label_SSP") +
  stat_compare_means(method = 'kruskal.test', label.y = 7) # Add global p-value
```





# Binary variables
We ran `t.test` between the below clinical variables and all sample scores. The test 
results were ordered based on p-value and top 6 of them are printed.

```{r}
vars <- c("msi", "summarylocation", "summarygrade", "summarystage")
```

### MSI
```{r echo=FALSE}
# MSI
var <- vars[1]
rm_ind <- which(is.na(data_all[,var]))
dat <- data_all[-rm_ind,]

msi.t.test <- sapply(seq_len(ncol(RAVmodel)), function(x) {
  res <- t.test(dat[,x] ~ dat[,var])
  p <- res$p.value
  return(p)
})

names(msi.t.test) <- paste0("RAV", seq_len(ncol(RAVmodel)))
msi.t.test <- msi.t.test[order(msi.t.test)]
# head(names(msi.t.test))
head(msi.t.test)
```

### Location
```{r echo=FALSE}
# Location
var <- vars[2]
rm_ind <- which(is.na(data_all[,var]))
dat <- data_all[-rm_ind,]

location.t.test <- sapply(seq_len(ncol(RAVmodel)), function(x) {
  res <- t.test(dat[,x] ~ dat[,var])
  p <- res$p.value
  return(p)
})

names(location.t.test) <- paste0("RAV", seq_len(ncol(RAVmodel)))
location.t.test <- location.t.test[order(location.t.test)]
# head(names(location.t.test))
head(location.t.test)
```

### Grade
```{r echo=FALSE}
# Grade
var <- vars[3]
rm_ind <- which(is.na(data_all[,var]))
dat <- data_all[-rm_ind,]

grade.t.test <- sapply(seq_len(ncol(RAVmodel)), function(x) {
  res <- t.test(dat[,x] ~ dat[,var])
  p <- res$p.value
  return(p)
})

names(grade.t.test) <- paste0("RAV", seq_len(ncol(RAVmodel)))
grade.t.test <- grade.t.test[order(grade.t.test)]
# head(names(grade.t.test))
head(grade.t.test)
```

### Stage
```{r echo=FALSE}
# Stage
var <- vars[4]
rm_ind <- which(is.na(data_all[,var]))
dat <- data_all[-rm_ind,]

stage.t.test <- sapply(seq_len(ncol(RAVmodel)), function(x) {
  res <- t.test(dat[,x] ~ dat[,var])
  p <- res$p.value
  return(p)
})

names(stage.t.test) <- paste0("RAV", seq_len(ncol(RAVmodel)))
stage.t.test <- stage.t.test[order(stage.t.test)]
# head(names(stage.t.test))
head(stage.t.test)
```

# Session Info
```{r}
sessionInfo()
```

