---
title: "Pathway Separation"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
abstract: "[**Source Code**](https://github.com/shbrief/GenomicSuperSignaturePaper/blob/master/Results/others/repeat_Fig5_Pathway_Separation.Rmd)"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>", message = FALSE, warning = FALSE, collapse = FALSE
)
```

# Setup
We benchmark one of the model validation measures, named as pathway separation,
from the previous [study](https://www.sciencedirect.com/science/article/pii/S240547121930119X?via%3Dihub)
(Figure 5). Briefly, pathway separation is defined as the ability of the 
signature model to keep non-overlapping signatures that can differentiate 
biologically similar pathways. 

## Load packages
```{r}
suppressPackageStartupMessages({
  library(GenomicSuperSignaturePaper)
  library(GenomicSuperSignature)
  library(dplyr)
})
```

## RAVmodel
To directly compare with the previous publication, we used the RAVmodel
annotated with the same priors: bloodCellMarkersIRISDMAP, svmMarkers, and 
canonicalPathways.

```{r message=FALSE, warning=FALSE, echo=FALSE}
## If GenomicSuperSignaturePaper is built locally with RAVmodel in inst/extdata
data.dir <- system.file("extdata", package = "GenomicSuperSignaturePaper")
RAVmodel <- readRDS(file.path(data.dir, "RAVmodel_PLIERpriors.rds"))
```

```{r eval=FALSE}
RAVmodel <- getModel("PLIERpriors", load=TRUE)
```

```{r}
RAVmodel

updateNote(RAVmodel)
```



# Pathway Separation
`cutoff_n` argument of `checkPathwaySeparation` function decides how many 
enriched pathways are used for the comparison. We tried both top 5 and top 1. 

## Type I and type II interferon
```{r}
ifn.alpha.set <- c("REACTOME_INTERFERON_ALPHA_BETA_SIGNALING")
ifn.gamma.set <- c("REACTOME_INTERFERON_GAMMA_SIGNALING")

checkPathwaySeparation(RAVmodel, ifn.alpha.set, ifn.gamma.set, 
                       cutoff_nes = NULL, cutoff_n = 5)
checkPathwaySeparation(RAVmodel, ifn.alpha.set, ifn.gamma.set, cutoff_n = 1)
```


## Myeloid lineage
Neutrophil vs. Monocyte

```{r}
neutrophil.set <- c("DMAP_GRAN3", "IRIS_Neutrophil-Resting", "SVM Neutrophils")
monocyte.set <- c("IRIS_Monocyte-Day0", "IRIS_Monocyte-Day1", 
                  "IRIS_Monocyte-Day7", "DMAP_MONO2", "SVM Monocytes",
                  "SVM Macrophages M0", "SVM Macrophages M1", 
                  "SVM Macrophages M2")

checkPathwaySeparation(RAVmodel, neutrophil.set, monocyte.set, 
                       cutoff_nes = NULL, cutoff_n = 5) 
checkPathwaySeparation(RAVmodel, neutrophil.set, monocyte.set, cutoff_n = 1)
```


## Proliferation
G1 vs. G2 cell cycle phases

```{r}
g1.set <- c("REACTOME_G1_S_TRANSITION", "REACTOME_M_G1_TRANSITION",
            "REACTOME_APC_C_CDH1_MEDIATED_DEGRADATION_OF_CDC20_AND_OTHER_APC_C_CDH1_TARGETED_PROTEINS_IN_LATE_MITOSIS_EARLY_G1", 
            "REACTOME_CYCLIN_E_ASSOCIATED_EVENTS_DURING_G1_S_TRANSITION_", 
            "REACTOME_G1_PHASE", "REACTOME_MITOTIC_M_M_G1_PHASES",
            "REACTOME_P53_DEPENDENT_G1_DNA_DAMAGE_RESPONSE", 
            "REACTOME_MITOTIC_G1_G1_S_PHASES", 
            "REACTOME_P53_INDEPENDENT_G1_S_DNA_DAMAGE_CHECKPOINT")
g2.set <- c("REACTOME_MITOTIC_G2_G2_M_PHASES", "REACTOME_G2_M_CHECKPOINTS")

checkPathwaySeparation(RAVmodel, g1.set, g2.set, 
                       cutoff_nes = NULL, cutoff_n = 5) 
checkPathwaySeparation(RAVmodel, g1.set, g2.set, cutoff_n = 1)
```


# Session Info
<details>
```{r}
sessionInfo()
```
</details>
