---
title: "CRC Subtyping using Sample Scores by RAVs"
author: "Sehyun Oh"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: false
    toc_depth: 3
  BiocStyle::pdf_document:
    toc: true
    toc_float: false
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, fig.align = "center",
                      out.width = "70%", fig.asp = 1)
```



# Setup
In this vignette, we perform colorectal cancer (CRC) subtyping analysis using
GenomicSuperSignature and benchmark it against Figure 4A of the [CRC paper (Ma *et al.*)](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-018-1511-4).  

## Load packages
```{r message=FALSE, warning=FALSE}
if (!"GenomicSuperSignaturePaper" %in% installed.packages())
    devtools::install_github("shbrief/GenomicSuperSignaturePaper")

suppressPackageStartupMessages({
  library(Biobase)
  library(tidyverse)
  library(GenomicSuperSignature)
  library(GenomicSuperSignaturePaper)
})
```

## RAVmodel
```{r message=FALSE, warning=FALSE, echo=FALSE}
## If GenomicSuperSignaturePaper is built locally with RAVmodel in inst/extdata
data.dir <- system.file("extdata", package = "GenomicSuperSignaturePaper")
RAVmodel <- readRDS(file.path(data.dir, "RAVmodel_C2.rds"))
```

```{r eval=FALSE}
RAVmodel <- getModel("C2", load=TRUE)
```

```{r}
RAVmodel

updateNote(RAVmodel)
```

## CRC validation datasets
CRC paper used colorectal cancer datasets from [curatedCRCData](https://bioconductor.org/packages/release/data/experiment/html/curatedCRCData.html) package. 
Eight of them are for training and ten are for validation.

```{r}
load("data/eSets/setNames.RData")
setNames

## Load validation samples
for (set in setNames) {
  load(paste0("data/eSets_new/", set, '.RData'))
}
```



# CRC subtyping with PCSSs
CRC paper actually used both training and validation datasets (total of 18) 
for Figure 4A. So here, we reproduce the same plot using all 18 datasets. 

Briefly, this plot summarizes the sample scores assigned to 3,567 tumor samples 
from 18 CRC studies. The samples in each of 18 datasets, assigned to either (i) 
one of the 4 previously proposed CMS subtypes by CRC Subtyping Consortium or 
(ii) not assigned to a CMS subtype (so 5 x 18 = 90 total groups), are 
represented by the mean (point) and standard deviation (error bar) of the 
sample scores. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
source("R/Fig4A_PCSS.R")
```

```{r echo=FALSE, eval=FALSE}
saveRDS(pA, "outputs/scatter_PCSS.rds")
```

If we use only 10 validation datasets:
```{r echo=FALSE}
## CRC paper actually used both training and validation datasets (total of 18) 
## for Figure 4. So here, we are using all of them. However, if you want to use 
## only 10 validation datasets, run the below chunk.

load("data/eSets/trainingSetNames.RData")
validationSetNames <- setdiff(setNames, trainingSetNames)
setNames <- validationSetNames
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
source("R/Fig4A_PCSS.R")
```

# CRC subtyping with RAVs
To evaluate the performance of RAVs compared to PCSSs, we searched for colon 
cancer associated RAVs in three different ways.      

1. We identified two RAVs, RAV1575 and RAV834, that have the highest Pearson 
correlation coefficients with PCSS1 and PCSS2, respectively (0.59 and 0.56). [(detail)](https://shbrief.github.io/GenomicSuperSignaturePaper/articles/CRC/PCSS_vs_RAV.html)    

2. We ran Kruskal-Wallis rank sum test between CMS subtypes and RAV-assigned 
scores. Two RAVs with the highest chi-square, RAV834 and RAV833, were selected 
for further testing. [(detail)](https://shbrief.github.io/GenomicSuperSignaturePaper/articles/CRC/metadata_based_search.html)    

3. We calculated validation scores for 18 colon cancer datasets used in CRC 
paper and collected top 10 validated RAVs from each dataset. We summarized the 
frequency of different RAVs validating each dataset without any additional 
filtering criteria and selected the top 2 most frequently validated RAVs, 
RAV188 and RAV832, which were captured 14 and 10 times, respectively. [(detail)](https://shbrief.github.io/GenomicSuperSignaturePaper/articles/CRC/CRC_Subtyping.html#highest-validation-score-1)   

In this section, we draw colon cancer subtype plots with these pairs of RAVs and
look up MeSH terms, enriched pathways, and related studies associated with them.




## Most similar to PCSS1/PCSS2
For plotting with RAVs, you can decided whether to use all 18 datasets or only 
10 validation datasets by setting `val_only` variable before running 
`R/Fig4A_plotting.R` script. Here, we are using only 10 validation datasets.

```{r}
val_only <- TRUE   # FALSE for 18 datasets, TRUE for 10 validation datasets
```

RAV1575 and RAV834 are identified as the RAVs most similar to PCSS1/2 in this [vignette](https://shbrief.github.io/GenomicSuperSignaturePaper/articles/CRC/PCSS_vs_RAV.html).

```{r message=FALSE, warning=FALSE}
sampleScore1 <- 1575
sampleScore2 <- 834
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

```{r echo=FALSE, eval=FALSE}
x <- source("R/Fig4A_plotting.R")
saveRDS(x, "outputs/scatter_1575_834.rds")
```

### Wordcloud
MeSH terms associated with `r paste0("RAV", sampleScore1, " (sampleScore1)")` 
and `r paste0("RAV", sampleScore2, " (sampleScore2)")`

```{r fig.width=9, fig.height=9}
drawWordcloud(RAVmodel, sampleScore1)
drawWordcloud(RAVmodel, sampleScore2)
```

### GSEA
Enriched pathways for `r paste0("RAV", sampleScore1, " (sampleScore1)")` 
and `r paste0("RAV", sampleScore2, " (sampleScore2)")`

```{r}
annotateRAV(RAVmodel, sampleScore1)
annotateRAV(RAVmodel, sampleScore2)
```

### Studies
Studies associated with `r paste0("RAV", sampleScore1, " (sampleScore1)")` 
and `r paste0("RAV", sampleScore2, " (sampleScore2)")`

```{r}
findStudiesInCluster(RAVmodel, sampleScore1, studyTitle = TRUE)
findStudiesInCluster(RAVmodel, sampleScore2, studyTitle = TRUE)
```


## Highest validation score

### Validation
Before applying `validate` function, any NA or Inf values in CRC datasets were
removed and the expression matrix was centered.

```{r message=FALSE, warning=FALSE}
load("data/eSets/setNames.RData")   # 18 CRC datasets
validated_ind_all <- vector(mode = "list", length = length(setNames))
names(validated_ind_all) <- setNames

for (set in setNames) {
  eSet <- get(set)
  exprs <- exprs(eSet) %>% rmNaInf
  exprs <- apply(exprs, 1, function(x) x - mean(x)) %>% t
  val_all <- validate(exprs, RAVmodel)
  validated_ind_all[[set]] <- validatedSignatures(val_all, num.out = 10, 
                                                  indexOnly = TRUE)
}
```

We combined the top validated RAVs from all 18 studies and checked the frequency
of them. Any RAV that appeared more often means they are consistently associated
with the common feature of CRC datasets. 

Top row is RAV number and the bottom row is the frequency of it making top 
validation. Because there were 18 datasets, the maximum frequency of showing 
up is 18. RAV188 and RAV832 are captured 14 and 10 times, respectively, as top
10 validated RAVs.

```{r}
## Top 10 validated
validated_ind_all %>% unlist %>% table %>% sort(., decreasing = TRUE)

## Top 5 validated
lapply(validated_ind_all, function(x) x[1:5]) %>%
  unlist %>% table %>% sort(., decreasing = TRUE)
```


### Top validated RAVs
```{r message=FALSE, warning=FALSE}
sampleScore1 <- 832
sampleScore2 <- 188
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

```{r echo=FALSE, eval=FALSE}
x <- source("R/Fig4A_plotting.R")
saveRDS(x, "outputs/scatter_832_188.rds")
```

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
sampleScore1 <- 832
sampleScore2 <- 833
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

### Wordcloud
MeSH terms associated with `r paste0("RAV", sampleScore1, " (sampleScore1)")` 
and `r paste0("RAV", sampleScore2, " (sampleScore2)")`

```{r fig.width=9, fig.height=9}
drawWordcloud(RAVmodel, sampleScore1)
drawWordcloud(RAVmodel, sampleScore2)
```

### GSEA
Enriched pathways for `r paste0("RAV", sampleScore1, " (sampleScore1)")` 
and `r paste0("RAV", sampleScore2, " (sampleScore2)")`

```{r}
annotateRAV(RAVmodel, sampleScore1)
annotateRAV(RAVmodel, sampleScore2)
```

### Studies
Studies associated with `r paste0("RAV", sampleScore1, " (sampleScore1)")` 
and `r paste0("RAV", sampleScore2, " (sampleScore2)")`

```{r}
findStudiesInCluster(RAVmodel, sampleScore1, studyTitle = TRUE)
findStudiesInCluster(RAVmodel, sampleScore2, studyTitle = TRUE)
```



## Best explaining metadata
We find the metadata-associated RAVs using Kruskal-Wallis Rank Sum Test. Further
detail can be found [here](https://shbrief.github.io/GenomicSuperSignaturePaper/articles/CRC/metadata_based_search.html).

```{r message=FALSE, warning=FALSE}
sampleScore1 <- 834
sampleScore2 <- 833
val_only <- TRUE   # FALSE for 18 datasets, TRUE for 10 validation datasets
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

```{r save_scatter_834_833, echo=FALSE, eval=FALSE}
x <- source("R/Fig4A_plotting.R")
saveRDS(x, "outputs/scatter_834_833.rds")

png("outputs/png/scatter_834_833.png", width = 400, height = 400)
print(x)$visible
dev.off()
```

### Wordcloud
MeSH terms associated with `r paste0("RAV", sampleScore1, " (sampleScore1)")` 
and `r paste0("RAV", sampleScore2, " (sampleScore2)")`

```{r fig.width=9, fig.height=9}
drawWordcloud(RAVmodel, sampleScore1)
drawWordcloud(RAVmodel, sampleScore2)
```

### GSEA
Enriched pathways for `r paste0("RAV", sampleScore1, " (sampleScore1)")` 
and `r paste0("RAV", sampleScore2, " (sampleScore2)")`

```{r}
annotateRAV(RAVmodel, sampleScore1)
annotateRAV(RAVmodel, sampleScore2)
```

### Studies
Studies associated with `r paste0("RAV", sampleScore1, " (sampleScore1)")` 
and `r paste0("RAV", sampleScore2, " (sampleScore2)")`

```{r}
findStudiesInCluster(RAVmodel, sampleScore1, studyTitle = TRUE)
findStudiesInCluster(RAVmodel, sampleScore2, studyTitle = TRUE)
```




# Manuscript Table
## Supplementary Table 5
This table contains the top 10 validated RAVs for 18 CRC datasets.

```{r}
CRC_validation_df <- data.frame(matrix(unlist(validated_ind_all), 
                                       nrow = length(validated_ind_all), 
                                       byrow = TRUE), 
                                stringsAsFactors=FALSE)
rownames(CRC_validation_df) <- names(validated_ind_all)
colnames(CRC_validation_df) <- paste0("validated_RAV_", 1:10)

CRC_validation_df[1:4, 1:4]
```

```{r echo=FALSE, eval=FALSE}
write.table(CRC_validation_df, "outputs/CRC_top10_validated_ind.tsv")
```


# Session Info

<details>
```{r}
sessionInfo()
```
</details>





```{r echo=FALSE, eval=FALSE}
## CMS by Candidate RAVs
dat <- readRDS("data/SummaryForFig4.rds")
score_ind <- grep("RAV", colnames(dat))

## Subset with 'cms_label_SSP'
dat <- dat %>%
  mutate(cms_label_SSP = cms_label_SSP %>%
           recode("unlabeled" = "not labeled"))

dat <- dat %>%
  group_by(study, cms_label_SSP) %>%
  dplyr::summarise(mean_cl1575 = mean(RAV1575),
                   mean_cl834 = mean(RAV834),
                   mean_cl833 = mean(RAV833),
                   sd_cl1575 = sd(RAV1575),
                   sd_cl834 = sd(RAV834),
                   sd_cl833 = sd(RAV833))

## Plot
cl_ind <- c(1575, 834, 833)
mean_names <- paste0("mean_cl", cl_ind)

for (i in seq_along(mean_names)) {
  ordered.dat <- dat[order(dat[,mean_names[i]]),]
  plot(dplyr::pull(ordered.dat, mean_names[i]),
       col = ordered.dat$cms_label_SSP,
       xlab = "", ylab = paste0("Score from RAV", cl_ind[i]))
}
```
