---
title: "Compare PCSS and RAVs"
author: "Sehyun Oh"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: false
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```


# Setup
## Load package
```{r}
suppressPackageStartupMessages({
  library(GenomicSuperSignature)
})
```

## PCSSs
[Ma *et al.*](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-018-1511-4) 
proposed a continuous scoring system (PCSS) based on the analysis of eight 
colorectal cancer microarray datasets comprising 1,867 samples, and found that
PCSSs are more closely correlated to microsatellite instability (MSI), grade, 
stage, and tumor location. 

```{r}
avgLoading <- read.table("data/avg_loadings.csv", 
                         header = TRUE, sep = ",")
avgLoading <- tibble::column_to_rownames(avgLoading, var="X")
colnames(avgLoading) <- paste0("PCSS", 1:4)

head(avgLoading)
```

## RAVmodel
```{r message=FALSE, warning=FALSE, echo=FALSE}
## If GenomicSuperSignaturePaper is built locally with RAVmodel in inst/extdata
data.dir <- system.file("extdata", package = "GenomicSuperSignaturePaper")
RAVmodel <- readRDS(file.path(data.dir, "RAVmodel_C2.rds"))
```

```{r eval=FALSE}
RAVmodel <- getModel("C2", load=TRUE)
```

```{r}
RAVmodel

updateNote(RAVmodel)
```

# RAVs vs. PCSSs
## Most similar to PCSS1/2
We identified RAV1575 and RAV834 as the most similar RAVs to PCSS1 and PCSS2, 
respectively, based on Pearson correlation coefficient.

```{r}
cg <- intersect(rownames(avgLoading), rownames(RAVmodel))
loading_cor <- abs(stats::cor(avgLoading[cg,], RAVindex(RAVmodel)[cg,], 
                              use="pairwise.complete.obs", method="pearson"))

max1 <- which.max(loading_cor[1,])  # max. correlation with PCSS1
max2 <- which.max(loading_cor[2,])  # max. correlation with PCSS2
```

```{r}
loading_cor[1, max1, drop = FALSE]
loading_cor[2, max2, drop = FALSE]
```


# Session Info

<details>
```{r}
sessionInfo()
```
</details>

