---
title: "Repeat CRC paper's Figure 4C"
author: "Sehyun Oh"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: false
    toc_depth: 3
  BiocStyle::pdf_document:
    toc: true
    toc_float: false
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message = FALSE)
```

# Setup
* CRC paper actually used both training and validation datasets (total of 18)
for Figure 4. `Fig4C_CMSvs.R` and `Fig4C_contScores.R` scripts are using only 10
validation datasets. If you want to use all 18 datasets, block the line 11-13 of
those scripts.   

```{r eval=FALSE}
## line 11-13
# load("data/eSets/trainingSetNames.RData")
# validationSetNames <- setdiff(setNames, trainingSetNames)
# setNames <- validationSetNames
```

* Fig4C_CMSvs.R script requires three environmental variables set before running:
`m2_name` (column name for the model to be compared to CMS), `m2_1` and `m2_2` (names
of the two model variables that form the group to be compared to CMS)   

* Fig4C_contScores.R script requires six environmental variables set before running:
`m1_name`, `m1_1`, `m1_2` for one model, `m2_name`, `m2_1`, `m2_2` for the other.


# Compare to CMS
## CMS vs. PCSS1/2
```{r fig.width=5, fig.height=3.5}
m2_name <- "PCSS"
m2_1 <- "PCSS1"
m2_2 <- "PCSS2"
source("R/Fig4C_CMSvs.R", print.eval = TRUE)
```

```{r echo=FALSE}
x <- source("R/Fig4C_CMSvs.R")
saveRDS(x, "outputs/boxplot_CMS_vs_PCSS.rds")
```


## CMS vs. RAV1575/834
RAV1575/834 are the most similar PCclsuters to PCSS1/2, respectively, based 
on Pearson correlation.

```{r fig.width=5, fig.height=3.5}
m2_name <- "RAV"
m2_1 <- "RAV1575"
m2_2 <- "RAV834"
source("R/Fig4C_CMSvs.R", print.eval = TRUE)
```

```{r echo=FALSE}
x <- source("R/Fig4C_CMSvs.R")
saveRDS(x, "outputs/boxplot_CMS_vs_1575_834.rds")
```

## CMS vs. RAV834/833
RAV834/833 have the largest r-squared score when we compared the samples
scores against the metadata, CMS.

```{r fig.width=5, fig.height=3.5}
m2_name <- "RAV"
m2_1 <- "RAV834"
m2_2 <- "RAV833"
source("R/Fig4C_CMSvs.R", print.eval = TRUE)
```

```{r echo=FALSE}
x <- source("R/Fig4C_CMSvs.R")
saveRDS(x, "outputs/boxplot_CMS_vs_834_833.rds")

png("outputs/png/boxplot_CMS_vs_834_833.png", width = 600, height = 500)
print(x)$visible
dev.off()
```



# Compare continuous scores
## PCSS vs. RAV1575/834
```{r fig.width=5, fig.height=3.5}
m1_name <- "PCSS"
m1_1 <- "PCSS1"
m1_2 <- "PCSS2"
m2_name <- "RAV"
m2_1 <- "RAV1575"
m2_2 <- "RAV834"

source("R/Fig4C_contScores.R", print.eval = TRUE)
```

```{r echo=FALSE}
x <- source("R/Fig4C_contScores.R")
saveRDS(x, "outputs/boxplot_PCSS_vs_1575_834.rds")
```

## PCSS vs. RAV834/833
```{r fig.width=5, fig.height=3.5}
m1_name <- "PCSS"
m1_1 <- "PCSS1"
m1_2 <- "PCSS2"
m2_name <- "RAV"
m2_1 <- "RAV834"
m2_2 <- "RAV833"

source("R/Fig4C_contScores.R", print.eval = TRUE)
```

```{r echo=FALSE}
x <- source("R/Fig4C_contScores.R")
saveRDS(x, "outputs/boxplot_PCSS_vs_834_833.rds")

png("outputs/png/boxplot_PCSS_vs_834_833.png", width = 600, height = 500)
print(x)$visible
dev.off()
```

## PCSS vs. RAV834/3290
RAV3290 is associated with "stage" metadata of CRC datasets.

```{r fig.width=5, fig.height=3.5}
m1_name <- "PCSS"
m1_1 <- "PCSS1"
m1_2 <- "PCSS2"
m2_name <- "RAV"
m2_1 <- "RAV834"
m2_2 <- "RAV3290"

source("R/Fig4C_contScores.R", print.eval = TRUE)
```

```{r echo=FALSE}
x <- source("R/Fig4C_contScores.R")
saveRDS(x, "outputs/boxplot_PCSS_vs_834_3290.rds")
```

## PCSS vs. RAV834/596
RAV596 is associated with "grade" metadata of CRC datasets.

```{r fig.width=5, fig.height=3.5}
m1_name <- "PCSS"
m1_1 <- "PCSS1"
m1_2 <- "PCSS2"
m2_name <- "RAV"
m2_1 <- "RAV834"
m2_2 <- "RAV596"

source("R/Fig4C_contScores.R", print.eval = TRUE)
```

```{r echo=FALSE}
x <- source("R/Fig4C_contScores.R")
saveRDS(x, "outputs/PCSS_vs_834_596.rds")
```

```{r}
sessionInfo()
```
