---
title: "Repeat CRC paper's Figure 4A"
author: "Sehyun Oh"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: false
    toc_depth: 3
  BiocStyle::pdf_document:
    toc: true
    toc_float: false
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Packages in GitHub
pkgs <- c("PCAGenomicSignatures", "PCAGenomicSignaturesPaper")
new_pkgs <- pkgs[!(pkgs %in% rownames(installed.packages()))]
if(length(new_pkgs) != 0) devtools::install_github(paste0("shbrief/", new_pkgs))

## BioC or CRAN packages
pkgs <- c("Biobase", "tidyverse")
new_pkgs <- pkgs[!(pkgs %in% rownames(installed.packages()))]
if(length(new_pkgs) != 0) BiocManager::install(new_pkgs)

## Load packages
suppressPackageStartupMessages({
  library(Biobase)
  library(tidyverse)
  library(PCAGenomicSignatures)
  library(PCAGenomicSignaturesPaper)
})
```

# Setup
In this vignette, we are reproducing Figure 4A of the [CRC paper](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-018-1511-4)
using PCAGenomicSignatures.

## PCAmodel
```{r message=FALSE, warning=FALSE, echo=FALSE}
data.dir <- system.file("extdata", package = "PCAGenomicSignaturesPaper")
PCAmodel <- readRDS(file.path(data.dir, "PCAmodel_C2.rds"))
PCAmodel
updateNote(PCAmodel)
```

## CRC validation datasets
```{r}
load("data/eSets/setNames.RData")
```

```{r echo=FALSE, eval=FALSE}
## CRC paper actually used both training and validation datasets (total of 18)
## for Figure 4. So here, we are using all of them. However, for Figure 4C, we'll
## try to use only 10 validation datasets. 
# load(file.path(wd, "trainingSetNames.RData"))
# validationSetNames <- setdiff(setNames, trainingSetNames)
```

```{r}
## Load validation samples
for (set in setNames) {
  load(paste0("data/eSets_new/", set, '.RData'))
}
```



# By PCSS1/PCSS2 
First, we simply reproduced the Figure 4A using PCSS1 and PCSS2.

```{r out.width="80%", out.height="80%", echo=FALSE, message=FALSE, warning=FALSE}
## dist’n of PCSS scores -------------------------------------------------------
df.results <- setNames %>% lapply(function(set) {
  eSet <- get(set)
  pdata <- pData(eSet)
  eSet_tmp <- eSet[, pdata$sample_type %in% "tumor"]
  exprs_tmp <- exprs(eSet_tmp)
  pdata_tmp <- pData(eSet_tmp)

  pdata_tmp$study <- set
  ind_rm <- grep("CRIS_", colnames(pdata_tmp))  # remove CIRS_* pData assigned to training datasets
  if (length(ind_rm) != 0) {pdata_tmp <- pdata_tmp[,-ind_rm]}
  return(pdata_tmp)
  
}) %>% Reduce('rbind', .)

## Subset with 'cms_label_SSP'
df.results <- df.results %>%
  mutate(cms_label_SSP = cms_label_SSP %>%
           recode("unlabeled" = "not labeled"))

df.results <- df.results %>%
  group_by(study, cms_label_SSP) %>%
  dplyr::summarise(mean_PCSS1 = mean(PCSS1),
                   mean_PCSS2 = mean(PCSS2),
                   sd_PCSS1 = sd(PCSS1),
                   sd_PCSS2 = sd(PCSS2))

## Plot Figure 4A
colors <- gg_color_hue(4)
colors.toplot <- c(colors, 'grey')
names(colors.toplot) <- c(paste0('CMS', 1:4), 'not labeled')

pA <- ggplot(df.results,
             aes(x = mean_PCSS1, y = mean_PCSS2, color = cms_label_SSP)) +
  geom_point() +
  geom_errorbar(aes(x = mean_PCSS1,
                    ymin = mean_PCSS2 - sd_PCSS2,
                    ymax = mean_PCSS2 + sd_PCSS2)) +
  geom_errorbarh(aes(y = mean_PCSS2,
                    xmin = mean_PCSS1 - sd_PCSS1,
                    xmax = mean_PCSS1 + sd_PCSS1)) +
  scale_color_manual(values = colors.toplot, name = "CMS Subtype") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  geom_hline(yintercept = 0, linetype = 'dashed') +
  geom_vline(xintercept = 0, linetype = 'dashed') +
  xlab('PCSS1') + ylab('PCSS2') +
  coord_cartesian(xlim = c(-2, 2.5)) +
  theme(legend.direction = "horizontal", legend.justification=c(0,1), legend.position=c(0,1),
        legend.background = element_rect(colour='black'))

print(pA)
```

```{r echo=FALSE}
saveRDS(pA, "outputs/scatter_PCSS.rds")
```

# By PCAGenomicSignatures
## Most similar to PCSS1/PCSS2
```{r message=FALSE, warning=FALSE, out.width="80%", out.height="80%"}
sampleScore1 <- 1575
sampleScore2 <- 834
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

```{r echo=FALSE}
x <- source("R/Fig4A_plotting.R")
saveRDS(x, "outputs/scatter_1575_834.rds")
```

### wordcloud
```{r out.width="60%", out.height="60%"}
drawWordcloud(PCAmodel, sampleScore1)
drawWordcloud(PCAmodel, sampleScore2)
```

### GSEA
```{r}
annotatePCcluster(PCAmodel, sampleScore1)
annotatePCcluster(PCAmodel, sampleScore2)
```


## Best validation scores overall on CRC validation datasets

### Validation
Before applying `validate` function, any NA or Inf values in CRC datasets were
removed and the expression matrix was centered.
 
```{r message=FALSE, warning=FALSE}
validated_ind_all <- vector(mode = "list", length = length(setNames))
names(validated_ind_all) <- setNames

for (set in setNames) {
  eSet <- get(set)
  exprs <- exprs(eSet) %>% rmNaInf
  exprs <- apply(exprs, 1, function(x) x - mean(x)) %>% t
  val_all <- validate(exprs, PCAmodel)
  validated_ind_all[[set]] <- validatedSignatures(val_all, num.out = 10, indexOnly = TRUE)
}
```

We combined the top validated RAVs from all 18 studies and checked the frequency 
of them. Any RAV that appeared more often means they are consistently associated
with the common feature of CRC datasets. 

Top row is RAV number and the bottom row is the frequency of it making top validation.
Because there were 18 datasets, the maximum frequeny of showing up is 18.

```{r}
## Top 10 validated
validated_ind_all %>% unlist %>% table %>% sort(., decreasing = TRUE)

## Top 5 validated
lapply(validated_ind_all, function(x) x[1:5]) %>%
  unlist %>% table %>% sort(., decreasing = TRUE)
```


### Top validated RAVs
```{r message=FALSE, warning=FALSE, out.width="80%", out.height="80%"}
sampleScore1 <- 832
sampleScore2 <- 188
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

```{r echo=FALSE}
x <- source("R/Fig4A_plotting.R")
saveRDS(x, "outputs/scatter_832_188.rds")

png("outputs/png/scatter_834_833.png", width = 400, height = 400)
print(x)$visible
dev.off()
```

```{r message=FALSE, warning=FALSE, out.width="80%", out.height="80%", echo=FALSE, eval=FALSE}
sampleScore1 <- 832
sampleScore2 <- 833
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

### wordcloud
```{r out.width="60%", out.height="60%"}
drawWordcloud(PCAmodel, sampleScore1)
drawWordcloud(PCAmodel, sampleScore2)
```

### GSEA
```{r}
annotatePCcluster(PCAmodel, sampleScore1)
annotatePCcluster(PCAmodel, sampleScore2)
```



## Identified using metadata, CMS
We find the metadata-associated RAVs using Kruskal-Wallis Rank Sum Test.

```{r message=FALSE, warning=FALSE, out.width="80%", out.height="80%"}
sampleScore1 <- 834
sampleScore2 <- 833
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

```{r echo=FALSE}
x <- source("R/Fig4A_plotting.R")
saveRDS(x, "outputs/scatter_834_833.rds")
```

### wordcloud
```{r out.width="60%", out.height="60%"}
drawWordcloud(PCAmodel, sampleScore1)
drawWordcloud(PCAmodel, sampleScore2)
```

### GSEA
```{r}
annotatePCcluster(PCAmodel, sampleScore1)
annotatePCcluster(PCAmodel, sampleScore2)
```





```{r fig.width=4, fig.height=4, echo=FALSE, eval=FALSE}
## CMS by Candidate PCclusters
dat <- readRDS("data/SummaryForFig4.rds")
score_ind <- grep("PCcluster", colnames(dat))

## Subset with 'cms_label_SSP'
dat <- dat %>%
  mutate(cms_label_SSP = cms_label_SSP %>%
           recode("unlabeled" = "not labeled"))

dat <- dat %>%
  group_by(study, cms_label_SSP) %>%
  dplyr::summarise(mean_cl1575 = mean(PCcluster1575),
                   mean_cl834 = mean(PCcluster834),
                   mean_cl833 = mean(PCcluster833),
                   sd_cl1575 = sd(PCcluster1575),
                   sd_cl834 = sd(PCcluster834),
                   sd_cl833 = sd(PCcluster833))

## Plot
cl_ind <- c(1575, 834, 833)
mean_names <- paste0("mean_cl", cl_ind)

for (i in seq_along(mean_names)) {
  ordered.dat <- dat[order(dat[,mean_names[i]]),]
  plot(dplyr::pull(ordered.dat, mean_names[i]),
       col = ordered.dat$cms_label_SSP,
       xlab = "", ylab = paste0("Score from PCcluster", cl_ind[i]))
}
```






```{r echo=FALSE}
## This chunk was executed only to save the summary table, not for vignettes.
CRC_validation_df <- data.frame(matrix(unlist(validated_ind_all), 
                                       nrow = length(validated_ind_all), 
                                       byrow = TRUE), 
                                stringsAsFactors=FALSE)
rownames(CRC_validation_df) <- names(validated_ind_all)
colnames(CRC_validation_df) <- paste0("val_ind_", 1:10)

write.table(CRC_validation_df, "outputs/CRC_top10_validated_ind.tsv")
```
