---
title: "Repeat CRC paper's Figure 4A"
author: "Sehyun Oh"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: false
    toc_depth: 3
  BiocStyle::pdf_document:
    toc: true
    toc_float: false
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Packages in GitHub
pkgs <- c("GenomicSuperSignature", "GenomicSuperSignaturePaper")
new_pkgs <- pkgs[!(pkgs %in% rownames(installed.packages()))]
if(length(new_pkgs) != 0) devtools::install_github(paste0("shbrief/", new_pkgs))

## BioC or CRAN packages
pkgs <- c("Biobase", "tidyverse")
new_pkgs <- pkgs[!(pkgs %in% rownames(installed.packages()))]
if(length(new_pkgs) != 0) BiocManager::install(new_pkgs)

## Load packages
suppressPackageStartupMessages({
  library(Biobase)
  library(tidyverse)
  library(GenomicSuperSignature)
  library(GenomicSuperSignaturePaper)
})
```

# Setup
In this vignette, we are reproducing Figure 4A of the [CRC paper](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-018-1511-4)
using GenomicSuperSignature.

## RAVmodel
```{r message=FALSE, warning=FALSE, echo=FALSE}
## If GenomicSuperSignaturePaper is built locally with RAVmodel in inst/extdata
data.dir <- system.file("extdata", package = "GenomicSuperSignaturePaper")
RAVmodel <- readRDS(file.path(data.dir, "RAVmodel_C2.rds"))
```

```{r eval=FALSE}
RAVmodel <- getModel("C2", load=TRUE)
```

```{r}
RAVmodel

updateNote(RAVmodel)
```

## CRC validation datasets
```{r}
load("data/eSets/setNames.RData")
```

CRC paper actually used both training and validation datasets (total of 18) for 
Figure 4. So here, we are using all of them. However, if you want to use only 10
validation datasets, run the below chunk.

```{r eval=FALSE}
# load("data/eSets/trainingSetNames.RData")
# validationSetNames <- setdiff(setNames, trainingSetNames)
# setNames <- validationSetNames
```

```{r}
## Load validation samples
for (set in setNames) {
  load(paste0("data/eSets_new/", set, '.RData'))
}
```



# Reproduce PCSS result
First, we simply reproduced the Figure 4A using PCSS1 and PCSS2.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=5.8, fig.height=5.5}
## dist’n of PCSS scores -------------------------------------------------------
df.results <- setNames %>% lapply(function(set) {
  eSet <- get(set)
  pdata <- pData(eSet)
  eSet_tmp <- eSet[, pdata$sample_type %in% "tumor"]
  exprs_tmp <- exprs(eSet_tmp)
  pdata_tmp <- pData(eSet_tmp)

  pdata_tmp$study <- set
  ind_rm <- grep("CRIS_", colnames(pdata_tmp))  # remove CIRS_* pData assigned to training datasets
  if (length(ind_rm) != 0) {pdata_tmp <- pdata_tmp[,-ind_rm]}
  return(pdata_tmp)
  
}) %>% Reduce('rbind', .)

## Subset with 'cms_label_SSP'
df.results <- df.results %>%
  mutate(cms_label_SSP = cms_label_SSP %>%
           recode("unlabeled" = "not labeled"))

df.results <- df.results %>%
  group_by(study, cms_label_SSP) %>%
  dplyr::summarise(mean_PCSS1 = mean(PCSS1),
                   mean_PCSS2 = mean(PCSS2),
                   sd_PCSS1 = sd(PCSS1),
                   sd_PCSS2 = sd(PCSS2))

## Plot Figure 4A
colors <- gg_color_hue(4)
colors.toplot <- c(colors, 'grey')
names(colors.toplot) <- c(paste0('CMS', 1:4), 'not labeled')

pA <- ggplot(df.results,
             aes(x = mean_PCSS1, y = mean_PCSS2, color = cms_label_SSP)) +
  geom_point() +
  geom_errorbar(aes(x = mean_PCSS1,
                    ymin = mean_PCSS2 - sd_PCSS2,
                    ymax = mean_PCSS2 + sd_PCSS2)) +
  geom_errorbarh(aes(y = mean_PCSS2,
                    xmin = mean_PCSS1 - sd_PCSS1,
                    xmax = mean_PCSS1 + sd_PCSS1)) +
  scale_color_manual(values = colors.toplot, name = "CMS Subtype") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  geom_hline(yintercept = 0, linetype = 'dashed') +
  geom_vline(xintercept = 0, linetype = 'dashed') +
  xlab('PCSS1') + ylab('PCSS2') +
  coord_cartesian(xlim = c(-2, 2.5)) +
  theme(legend.direction = "horizontal", legend.justification=c(0,1), legend.position=c(0,1),
        legend.background = element_rect(colour='black'))

print(pA)
```

```{r echo=FALSE, eval=FALSE}
saveRDS(pA, "outputs/scatter_PCSS.rds")
```

# CRC subtyping using GenomicSuperSignature
## Most similar to PCSS1/PCSS2

[Update on 02.18.21] To use only 10 CRC validation datasets, you set the `val_only` 
variable before running `R/Fig4A_plotting.R` script.

```{r message=FALSE, warning=FALSE, fig.width=5.8, fig.height=5.5}
sampleScore1 <- 1575
sampleScore2 <- 834
val_only <- TRUE   # FALSE for 18 datasets, TRUE for 10 validation datasets
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

```{r echo=FALSE, eval=FALSE}
x <- source("R/Fig4A_plotting.R")
saveRDS(x, "outputs/scatter_1575_834.rds")
```

### wordcloud
```{r fig.width=7, fig.height=7}
drawWordcloud(RAVmodel, sampleScore1)
drawWordcloud(RAVmodel, sampleScore2)
```

### GSEA
```{r}
annotateRAV(RAVmodel, sampleScore1)
annotateRAV(RAVmodel, sampleScore2)
```


## Best validation scores overall on CRC validation datasets

### Validation
Before applying `validate` function, any NA or Inf values in CRC datasets were
removed and the expression matrix was centered.
 
```{r message=FALSE, warning=FALSE}
validated_ind_all <- vector(mode = "list", length = length(setNames))
names(validated_ind_all) <- setNames

for (set in setNames) {
  eSet <- get(set)
  exprs <- exprs(eSet) %>% rmNaInf
  exprs <- apply(exprs, 1, function(x) x - mean(x)) %>% t
  val_all <- validate(exprs, RAVmodel)
  validated_ind_all[[set]] <- validatedSignatures(val_all, num.out = 10, indexOnly = TRUE)
}
```

We combined the top validated RAVs from all 18 studies and checked the frequency 
of them. Any RAV that appeared more often means they are consistently associated
with the common feature of CRC datasets. 

Top row is RAV number and the bottom row is the frequency of it making top validation.
Because there were 18 datasets, the maximum frequeny of showing up is 18.

```{r}
## Top 10 validated
validated_ind_all %>% unlist %>% table %>% sort(., decreasing = TRUE)

## Top 5 validated
lapply(validated_ind_all, function(x) x[1:5]) %>%
  unlist %>% table %>% sort(., decreasing = TRUE)
```


### Top validated RAVs
```{r message=FALSE, warning=FALSE, fig.width=5.8, fig.height=5.5}
sampleScore1 <- 832
sampleScore2 <- 188
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

```{r echo=FALSE, eval=FALSE}
x <- source("R/Fig4A_plotting.R")
saveRDS(x, "outputs/scatter_832_188.rds")
```

```{r message=FALSE, warning=FALSE, out.width="80%", out.height="80%", echo=FALSE, eval=FALSE}
sampleScore1 <- 832
sampleScore2 <- 833
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

### wordcloud
```{r fig.width=7, fig.height=7}
drawWordcloud(RAVmodel, sampleScore1)
drawWordcloud(RAVmodel, sampleScore2)
```

### GSEA
```{r}
annotateRAV(RAVmodel, sampleScore1)
annotateRAV(RAVmodel, sampleScore2)
```



## Identified using metadata
We find the metadata-associated RAVs using Kruskal-Wallis Rank Sum Test. Further
detail can be found [here](https://shbrief.github.io/GenomicSuperSignaturePaper/articles/CRC/metadata_based_search.html).

```{r message=FALSE, warning=FALSE, fig.width=5.8, fig.height=5.5}
sampleScore1 <- 834
sampleScore2 <- 833
val_only <- TRUE   # FALSE for 18 datasets, TRUE for 10 validation datasets
source("R/Fig4A_plotting.R", print.eval = TRUE)
```

```{r save_scatter_834_833, echo=FALSE, eval=FALSE}
x <- source("R/Fig4A_plotting.R")
saveRDS(x, "outputs/scatter_834_833.rds")

png("outputs/png/scatter_834_833.png", width = 400, height = 400)
print(x)$visible
dev.off()
```

### wordcloud
```{r fig.width=6.5, fig.height=6.5}
drawWordcloud(RAVmodel, sampleScore1)
drawWordcloud(RAVmodel, sampleScore2)
```

### GSEA
```{r}
annotateRAV(RAVmodel, sampleScore1)
annotateRAV(RAVmodel, sampleScore2)
```

# Session Info
```{r}
sessionInfo()
```




```{r fig.width=4, fig.height=4, echo=FALSE, eval=FALSE}
## CMS by Candidate RAVs
dat <- readRDS("data/SummaryForFig4.rds")
score_ind <- grep("RAV", colnames(dat))

## Subset with 'cms_label_SSP'
dat <- dat %>%
  mutate(cms_label_SSP = cms_label_SSP %>%
           recode("unlabeled" = "not labeled"))

dat <- dat %>%
  group_by(study, cms_label_SSP) %>%
  dplyr::summarise(mean_cl1575 = mean(RAV1575),
                   mean_cl834 = mean(RAV834),
                   mean_cl833 = mean(RAV833),
                   sd_cl1575 = sd(RAV1575),
                   sd_cl834 = sd(RAV834),
                   sd_cl833 = sd(RAV833))

## Plot
cl_ind <- c(1575, 834, 833)
mean_names <- paste0("mean_cl", cl_ind)

for (i in seq_along(mean_names)) {
  ordered.dat <- dat[order(dat[,mean_names[i]]),]
  plot(dplyr::pull(ordered.dat, mean_names[i]),
       col = ordered.dat$cms_label_SSP,
       xlab = "", ylab = paste0("Score from RAV", cl_ind[i]))
}
```






```{r echo=FALSE}
## This chunk was executed only to save the summary table, not for vignettes.
CRC_validation_df <- data.frame(matrix(unlist(validated_ind_all), 
                                       nrow = length(validated_ind_all), 
                                       byrow = TRUE), 
                                stringsAsFactors=FALSE)
rownames(CRC_validation_df) <- names(validated_ind_all)
colnames(CRC_validation_df) <- paste0("val_ind_", 1:10)

write.table(CRC_validation_df, "outputs/CRC_top10_validated_ind.tsv")
```
