---
title: "Analyze 5 TCGA datasets using RAVmodel"
author: "Sehyun Oh"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: false
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{5 TCGA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message = FALSE)
```

# Setup
## Load packages
```{r}
suppressPackageStartupMessages({
  library(GenomicSuperSignature)
  library(dplyr)
  library(ggplot2)
})
```

## TCGA datasets
`TCGA_validationDatasets` is a list containing 6 TCGA datasets: COAD, BRCA, LUAD,
READ, UCEC, and OV. First 5 are raw read counts from `GSEABenchmarkeR` package with
log2(count + 1) transformation. Also, `GSEABenchmarkeR::loadEData` excluded genes with 
cpm < 2 in more than half of the samples. TCGA-OV dataset is from `curatedOvarianData`
package. This training dataset file was build using `R/build_TCGA_validationDatasets.R`
script.

```{r collapse=TRUE}
load("data/TCGA_validationDatasets.rda")
names(TCGA_validationDatasets)

# not using OV for this vigenette
datasets <- TCGA_validationDatasets[1:5]
```

## RAVmodel
This RAVmodel is built using top 20 PCs of 536 studies. The number of cluster was
decided by `d=2.25` and MSigDB C2 gene sets are used for GSEA-based annotation.

```{r message=FALSE, warning=FALSE, echo=FALSE}
## If GenomicSuperSignaturePaper is built locally with RAVmodel in inst/extdata
data.dir <- system.file("extdata", package = "GenomicSuperSignaturePaper")
RAVmodel <- readRDS(file.path(data.dir, "RAVmodel_C2.rds"))
```

```{r eval=FALSE}
RAVmodel <- getModel("C2", load=TRUE)
```

```{r}
RAVmodel

updateNote(RAVmodel)
```




# Multi-comparison (only TCGA)
## heatmapTable (all)
Here, we plot the heatmapTable of validation result from multiple studies. By matching
this result with metadata, you can identify RAVs that are more strongly associated 
with specific features. 

```{r fig.width=5.5, fig.height=4, out.width="70%", out.height="70%"}
# This process takes little time due to the size of datasets.
val_all <- validate(datasets, RAVmodel)
heatmapTable(val_all) 
```

It seems like RAV221 and RAV868 are specific to BRCA while RAV832 is strongly 
associated with colon/rectal cancers. 

## heatmapTable (one dataset)
If you provide validation result from one dataset, `heatmapTable` include the average
silhouette width as a reference.

### COAD
```{r fig.width=3.5, fig.height=2, out.width="70%", out.height="70%"}
val_coad <- validate(datasets[["COAD"]], RAVmodel)
heatmapTable(val_coad) 
```

### READ
```{r fig.width=3.5, fig.height=2, out.width="70%", out.height="70%"}
val_read <- validate(datasets[["READ"]], RAVmodel)
heatmapTable(val_read) 
```

### BRCA
RAV221 shows the highest validation score with a positive avgerage silhouette width. 
```{r fig.width=3.5, fig.height=2, out.width="70%", out.height="70%"}
val_brca <- validate(datasets[["BRCA"]], RAVmodel)
heatmapTable(val_brca)
```

RAV868 didn't score top 5, so we checked a couple more top validated RAVs. RAV868
is ranked 6th based on the score with negative average silhouette width.
``````{r fig.width=4, fig.height=2, out.width="70%", out.height="70%"}
heatmapTable(val_brca, num.out = 7) 
```

## MeSH terms and associated studies
### BRCA-associated
RAV221 and RAV868 are specific to BRCA.

#### RAV221
RAV221 consists of three breast cancer studies (ERP016798, SRP023262, and SRP11343)
and top 10 enriched pathways are associated with breast cancer. 
```{r fig.height=5, fig.width=5, out.width="70%", out.height="70%"}
ind <- 221
findStudiesInCluster(RAVmodel, ind)
subsetEnrichedPathways(RAVmodel, ind) %>% as.data.frame
drawWordcloud(RAVmodel, ind)
```

#### RAV868
```{r fig.height=5, fig.width=5, out.width="70%", out.height="70%"}
ind <- 868
findStudiesInCluster(RAVmodel, ind)
subsetEnrichedPathways(RAVmodel, ind) %>% as.data.frame
drawWordcloud(RAVmodel, ind)
```

### COAD/READ-associated
#### RAV832
```{r fig.height=5, fig.width=5, out.width="70%", out.height="70%"}
ind <- 832
findStudiesInCluster(RAVmodel, ind)
subsetEnrichedPathways(RAVmodel, ind) %>% as.data.frame
drawWordcloud(RAVmodel, ind)
```



# Multi-comparison (TCGA + others)
## heatmapTable (all)
Here, we added SLE-WB microarray dataset and 4 colon cancer microarray dataasets
to 5 TCGA dataset and `scoreCutoff` is set to 0.68 instead of the default 0.7.

```{r echo=FALSE, warning=FALSE, message=FALSE}
new_datasets <- lapply(datasets, assay)

## SLE-WB dataset
fname <- "~/data2/GenomicSuperSignaturePaper/Results/SLE-WB/data/SLE_WB_all_microarray_QN_zto_before_with_GeneSymbol.pcl"
exprs <- readr::read_tsv(fname) %>% as.data.frame
rownames(exprs) <- exprs$GeneSymbol
sle_wb <- as.matrix(exprs[,3:ncol(exprs)])
new_datasets$SLE <- sle_wb

## CRC dataset
load("~/data2/GenomicSuperSignaturePaper/Results/CRC/data/eSets_new/GSE14095_eset.RData")
load("~/data2/GenomicSuperSignaturePaper/Results/CRC/data/eSets_new/GSE17536_eset.RData")
load("~/data2/GenomicSuperSignaturePaper/Results/CRC/data/eSets_new/GSE2109_eset.RData")
load("~/data2/GenomicSuperSignaturePaper/Results/CRC/data/eSets_new/GSE39582_eset.RData")
new_datasets$GSE14095 <- log2(exprs(GSE14095_eset)+1)
new_datasets$GSE17536 <- log2(exprs(GSE17536_eset)+1)
new_datasets$GSE2109 <- log2(exprs(GSE2109_eset)+1)
new_datasets$GSE39582 <- log2(exprs(GSE39582_eset)+1)
```

```{r}
names(new_datasets)
```

```{r echo=FALSE, out.width="90%", out.height="90%"}
new_val_all <- validate(new_datasets, RAVmodel)
heatmapTable(new_val_all, scoreCutoff = 0.68) 
```

Based on this multi-datasets validation table,       
- RAV23 and RAV1551 are SLE-specific    
- RAV188 seems to be COAD-specific, different from RAV832 which is associated with both COAD and READ   


## MeSH terms and associated studies
### SLE-associated
#### RAV23
```{r fig.height=7, fig.width=7, out.width="70%", out.height="70%"}
ind <- 23
findStudiesInCluster(RAVmodel, ind)
subsetEnrichedPathways(RAVmodel, ind) %>% as.data.frame
drawWordcloud(RAVmodel, ind)
```

#### RAV1551
```{r fig.height=7, fig.width=7, out.width="70%", out.height="70%"}
ind <- 1551
findStudiesInCluster(RAVmodel, ind)
subsetEnrichedPathways(RAVmodel, ind) %>% as.data.frame
drawWordcloud(RAVmodel, ind)
```

### COAD-specific
```{r fig.height=7, fig.width=7, out.width="70%", out.height="70%"}
ind <- 188
findStudiesInCluster(RAVmodel, ind)
subsetEnrichedPathways(RAVmodel, ind, both=TRUE) %>% as.data.frame
drawWordcloud(RAVmodel, ind)
```



# Manuscript Figures
## Fig 2. Connect new data with the existing databases
A RAVmodel provides a rich resource for understanding new or user-supplied datasets in the context of all other RAVmodel datasets. A) Validation of multiple datasets. Each dataset was subjected to PCA and correlated to all possible RAVs using Pearson correlation. RAVs with correlation above 0.7 in at least one dataset were included. RAV221 and RAV868 seem to be associated with breast cancer while RAV832 is associated with colon and rectal cancer. B) Validation of a single dataset, TCGA-BRCA. Top 5 validated RAVs (score, bottom panel) and their average silhouette width (avg.sw, top panel) are shown. C) MeSH terms associated with RAV221 can aid in interpreting the biological context of the RAV when plotted as a word cloud. D) Studies contributing to RAV221. Study accession number (studyName column) and the title of a study (title column) are shown here. E) Top 10 enriched pathways in RAV221.

<br>

```{r manuscript_fig2, fig.width=11, fig.height=8, echo=FALSE, eval=FALSE}
## Draw Fig2A/2B heatmapTable 
a <- grid::grid.grabExpr(ComplexHeatmap::draw(heatmapTable(val_all)))
b <- grid::grid.grabExpr(ComplexHeatmap::draw(heatmapTable(val_brca)))

## Save Fig.2C wordcloud
png("manuscript_Fig2C.png", width = 350, height = 350)
drawWordcloud(RAVmodel, ind = 221)
dev.off()
c <- png::readPNG("manuscript_Fig2C.png") %>% grid::rasterGrob() %>% gridExtra::grid.arrange()

## Make Fig2D/2E tables
library(ggpubr)
ind <- 221
fig2d <- findStudiesInCluster(RAVmodel, ind)
fig2e <- subsetEnrichedPathways(RAVmodel, ind) %>% as.data.frame

library(flextable)
d <- flextable(fig2d) %>% width(., j = 2, width = 4) 
d <- grid::rasterGrob(as_raster(d))

e <- flextable(fig2e) %>% width(., j = 1, width = 5.3) 
e <- grid::rasterGrob(as_raster(e))

# ## If I want to add rownames
# fig2e <- subsetEnrichedPathways(RAVmodel, ind) %>% 
#   as.data.frame %>% 
#   tibble::rownames_to_column(., " ")
# e <- flextable(fig2e) %>% width(., j = 2, width = 4) 
# e <- grid::rasterGrob(as_raster(e))

## Make the complete Figure 2
ac <- ggarrange(a, c, 
                labels = c("A", "C"), vjust = c(2, 6), hjust = -1.5,
                nrow = 2, 
                heights = c(0.5, 1),
                font.label = list(size = 15))
bde <- ggarrange(b, d, e, 
                 labels = c("B", "D", "E"), vjust = c(6,1,1), hjust = c(-5,0,0),
                 nrow = 3, align = "v",
                 heights = c(0.8, 0.45, 0.8),
                 font.label = list(size = 15))

Fig2_all <- ggarrange(ac, bde, ncol = 2, widths = c(1, 1.2), align = "hv") 
Fig2_all

## Save the complete Figure 2
png("manuscript_Fig2.png", width = 715, height = 520)
Fig2_all
dev.off()
```

```{r fig.width=11, fig.height=8, echo=FALSE}
library(EBImage)
img <- readImage("manuscript_Fig2.png")
display(img, method = "raster")
```

```{r fig2_ver1, fig.width=12, fig.height=10, echo=FALSE, eval=FALSE}
# ## Make the complete Figure 2
# ab <- ggarrange(a, b, ncol = 2, labels = c("A", "B"), vjust = 9, hjust = -7,
#                 widths = c(1, 0.8), font.label = list(size = 15))
# cde <- ggarrange(ggarrange(c, labels = "C", vjust = 3, hjust = -7),  # 2C wordcloud
#                  ggarrange(d, e,  # 2D and 2E tables  
#                            labels = c("D", "E"), vjust = 1, hjust = 0,
#                            nrow = 2, align = "hv",
#                            heights = c(0.8, 1.4)), 
#                  ncol = 2,
#                  widths = c(1, 1.2),
#                  font.label = list(size = 15),
#                  align = "hv")
# Fig2_all <- ggarrange(ab, cde, nrow = 2)   

# ## Save the complete Figure 2
# png("Fig2.png", width = 1200, height = 1000)
# Fig2_all
# dev.off()
```




## Sup.Fig 4. Colon and rectal cancer associated RAV
Based on Figure 2A, RAV832 seems to be associated with TCGA-COAD and TCGA-READ. Top validation results of A) TCGA-COAD and B) TCGA-READ include RAV832 with the negative average silhouette width. C) MeSH terms associated with RAV832. D) Studies contributing to RAV832. E) MSigDB C2 gene sets enriched in RAV832.

<br>

```{r manuscript_sup_fig4, fig.width=11, fig.height=8, echo=FALSE, eval=FALSE}
## Draw Sup.Fig4A/4B heatmapTable 
a <- grid::grid.grabExpr(ComplexHeatmap::draw(heatmapTable(val_coad)))
b <- grid::grid.grabExpr(ComplexHeatmap::draw(heatmapTable(val_read)))

## Save Sup.Fig.4C wordcloud
png("manuscript_Sup_Fig4C.png", width = 350, height = 350)
drawWordcloud(RAVmodel, ind = 832)
dev.off()
c <- png::readPNG("manuscript_Sup_Fig4C.png") %>% grid::rasterGrob() %>% gridExtra::grid.arrange()

## Make Sup.Fig4D/4E tables
library(ggpubr)
ind <- 832
fig2d <- findStudiesInCluster(RAVmodel, ind)
fig2e <- subsetEnrichedPathways(RAVmodel, ind) %>% as.data.frame

library(flextable)
d <- flextable(fig2d) %>% width(., j = 2, width = 6) 
d <- grid::rasterGrob(as_raster(d))

e <- flextable(fig2e) %>% width(., j = 1, width = 5.3) 
e <- grid::rasterGrob(as_raster(e))
```

```{r sup_fig4_ver1, fig.width=9.5, fig.height=6.5, echo=FALSE, eval=FALSE}
## Make the complete Sup.Figure 4
abc <- ggarrange(a, b, c, 
                 nrow = 3, heights = c(0.7, 0.7, 1.4), 
                 labels = c("A", "B", "C"), vjust = 1.5, hjust = -2,
                 font.label = list(size = 15))
de <- ggarrange(d, e, 
                nrow = 2, heights = c(0.8, 1.4),
                labels = c("D", "E"), vjust = c(4.5, 7), hjust = 2.5,
                align = "v",
                font.label = list(size = 15))

Sup_Fig4_all <- ggarrange(abc, de, ncol = 2, align = "h")
Sup_Fig4_all

# ## Save the complete Sup.Figure 4
# ##### Worked better to save it as Rmd display
# png("manuscript_Sup_Fig4.png", width = 950, height = 650)
# Sup_Fig4_all
# dev.off()
```

```{r fig.width=9.5, fig.height=6.5, echo=FALSE}
library(EBImage)
img <- readImage("manuscript_Sup_Fig4.png")
display(img, method = "raster")
```

```{r sup_fig4_ver2, fig.width=12, fig.height=10, echo=FALSE, eval=FALSE}
## Make the complete Sup.Figure 4
abc <- ggarrange(a, b, c, 
                 ncol = 3, widths = c(1, 1, 1), 
                 labels = c("A", "B", "C"), vjust = c(9,9,6), hjust = -1,
                 font.label = list(size = 15))
de <- ggarrange(d, e, 
                nrow = 2, heights = c(0.8, 1.4),
                labels = c("D", "E"), vjust = 1.5, hjust = -15,
                align = "v",
                font.label = list(size = 15))

Sup_Fig4_all <- ggarrange(abc, de, nrow = 2)

## Save the complete Sup.Figure 4
png("manuscript_Sup_Fig4.png", width = 1200, height = 1000)
Sup_Fig4_all
dev.off()
```


# Session Info

<details>
    ```{r}
sessionInfo()
```
</details>
    
